Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-10 content/contribution/continuous-datalog.tex
Incremental changes to datalog description

\begin{nestedsection}{Continuous Datalog}{semantics}
The first requirement of stream reasoning is a well-defined semantics.
This must take into account the continuous nature of a subset of the
base axioms of a stream reasoning problem, as well as its
instantaneous truth and entailments, in line with the work of the
reasoning community.  However, it must also provide a continuous
interpretation of those instantaneous entailments in order to
integrate with other stream-processing technologies.  More
specifically, an appropriate semantics for stream reasoning on the
Semantic Web must: allow the expression of OWL 2 ontologies to some
degree of expressivity; not sacrifice the precise semantics thereof
whilst incorporating the semantics of stream processing; and allow for
the continuous derivation of entailments in a concise manner.

	Continuous Datalog is the extension of positive Datalog with stream-processing semantics, with a view to accommodating the aims of the latter within the well-defined proof-theoretic semantics of the former, which are shared with RIF-Core and therefore support reasoning at the expressivity of OWL 2 RL.
	This comprises the extension of Datalog programs to include sources of continuously generated/reported facts, and the application of sliding window semantics over those sources, thereby providing for the interpretation of Continuous Datalog programs as sequences of instantaneous states that are each valid positive Datalog programs.
	As such, these states each inherit the semantics of positive Datalog.
	Furthermore, it involves the continuous derivation of entailments from this sequence of instantaneous states as a set of streams of facts;
	these streams share a data model with the streams of continuous queries, as defined by the stream processing community, thereby allowing the integration of Continuous Datalog programs into stream processing pipelines.

@@

\begin{definition}[Datalog Program]
\labeldef{continuous datalog: datalog program}
A conventional Datalog program $\Pi = \langle R, F\rangle$ is a tuple
which comprises a finite set of rules $R$ and a finite set of facts
$F$. Facts are extensional atomic formulae of first-order logic, and
rules are of the form:
\[\alpha_0 \text{ :- } \alpha_1 , \ldots , \alpha_n\]
where $\alpha_0$ is an intensional atomic formula and
$\alpha_1 \ldots \alpha_n$ are a mixture of intensional and
extensional atomic formulae.

@@

When considering a time-annotated Datalog program, in which there may be
many distinct instances of a given fact, temporal information regarding when
instances of that fact became true would be lost if such a program were to entail
sets of plain facts. As such, where a fact $f$ would be entailed by some Datalog
program ${\Pi = \langle R, \{ strip(x) : x \in W^{S,x}_i \} \rangle}$,
for some window over some stream $S$, an instance ${f | t}$ of $f$ should be
entailed by the time-annotated Datalog program
${\Pi^T = \langle R, W^{S,x}_i \rangle \vdash f | t}$.

@@

It should be noted that \refdef{continuous datalog: justification} of entailment
justification is equally applicable to time-annotated Datalog programs,
simply replacing entailed and justifying facts with fact instances.
Furthermore, each justification for $f$ in some $F^T$ entails a unique
instance ${f | t}$, such that many instances of $f$ may occur in the set
$E^T$ of fact instances entailed by
${\Pi^T = \langle R, W^{S,x}_i \rangle}$, each with a single
justification.  As such, entailment justification for fact instances may be
defined formally by \refdef{continuous datalog: instance justification}.
\begin{definition}[Instance Entailment Justification]
\labeldef{continuous datalog: instance justification}
	A justification for some entailment ${f|t}$ of some time-annotated Datalog program
	${\Pi^T = \langle R , F^T \rangle \vdash f|t}$ is the single minimal set ${J^T \subseteq F^T}$
	such that ${\Pi^T' = \langle R , J^T \rangle \vdash f|t}$.
	The mapping of the fact instance ${f | t}$ to the minimal subset $J^T$
	within ${\Pi^T = \langle R, F^T \rangle}$ that justifies it is expressed
	as the function ${just : \mathbb{\Pi^T} , \mathbb{F^T} \rightarrow \{ \mathbb{F^T} \}}$.
\end{definition}

@@

\begin{definition}[Continuous Datalog Program]\label{def:continuous datalog: CDP}
A Continuous Datalog Program $\Pi^C$ is a tuple ${\langle R, F, S, r
\rangle}$ which comprises a finite set of rules $R$, a finite set of persistent
facts $F$, a stream of facts $S$ and a range ${r \in \mathbb{N}}$ for
the sliding window over $S$. This specifies that the streamed instances from $S$ are
to be considered valid from their timestamp for the duration represented
by $r$.

@@

\begin{definition}[State of a $\Pi^C$]\label{def:continuous datalog: CDPt}
The state of a $\Pi^C$ at a given instant $i$ is the time-annotated
Datalog program ${\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle}$,
where ${F^0 = \{ f | 0 : f \in F \}}$ is the set of persistent fact instances produced by
annotating each fact in $f$ in $F$ as becoming true at the first recognised instant.

As both $F$ and ${\{ strip(x) : x \in W^{S,r}_{i} \}}$ are sets of conventional
Datalog facts, the tuple ${\langle R, F \cup \{ strip(x) : x \in W^{S,r}_{i} \}}$
constitutes a valid Datalog program that may be evaluated by the truth
function of Datalog $T^{D}$, the result of which is the definition for the
instantaneous truth of $\Pi^C$ at instant $i$ by $T^{CD}$.
\begin{multline*}
T^{CD} \left( \langle R, F, S, r \rangle, i \right) = \\
	T^{D} \left( \langle R, F \cup \{ strip(x) : x \in W^{S,r}_{i} \} \right)
\end{multline*}

The instantaneous entailments of a $\Pi^C_i$ are the union of the set of
persistent entailments of ${\langle R, F^0 \rangle \vDash E^0}$ (which are
guaranteed to be entailed by every state, as positive Datalog is monotonic),
and the set $E^T_{i}$ of transient entailments whose justifications are at
least partially contained in $W^{S,r}_i$.
\begin{equation*}
\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle \vDash E^0 \cup E^T_{i}
\end{equation*}
$E^T_i$ is defined more formally as
${E^T_{i} = \{ x : \left( \Pi^C_i \vdash x \right) \wedge
\left( just(\Pi^C_i,x) \cap W^{S,r}_i \neq \varnothing \right) \}}$.
By this definition and Axiom~\ref{axiom:continuous datalog: entailment disjointness},
$E^0$ is guaranteed to be disjoint from every $E^T_i$
for every state $\Pi^C_i$ of every Continuous Datalog Program $\Pi^C$.
\end{definition}

@@

From this, we can see that each entailed fact instance ${f|t}$ in some
$E^T_{i}$, of the form ${f|t = \langle s, p, o \rangle | t}$, has a timestamp $t$ that is
the real-time associated with the state $\Pi^C_{i_{e}}$ in which its particular justification for
${f|t}$ is first complete.
As such, the stream of entailed fact instances $P$ of a $\Pi^C$ takes precisely
the same form as the streams of facts already described, thus
producing a concise stream of entailments that may be read by
further stream processing.  In addition, the stream of negations $N$
of a $\Pi^C$ also takes this form, but differs from the streams
discussed thus far in that it is not ordered by the entailment/assertion time
of the entailed fact instances, but by the real-time associated with
the first state $\Pi^C_{i_{n}}$ in which their
justification \emph{ceases} to be complete,
i.e. ${just(\Pi^C_{i_{e}}, f|t) \subseteq W^{S,r}_{i_n - 1}}$ but
${just(\Pi^C_{i_{e}}, f|t) \not\subseteq W^{S,r}_{i_n}}$.

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.  However, it can be shown from these axioms, as in
\ref{sec:semantics: proof for entailments}, that it is also possible
to reconstruct exactly the instantaneous set $E^T_{i}$, of instances
entailed at any given instant $i$ by any $\Pi^C = \{ R, F, S, r \}$,
from the sliding windows of sufficient range over its entailed streams
$P$ and $N$ at $i$:
\[ E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i} \]
where ${y \geq x \geq r}$. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^C_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{correct} set $E^T_{i}$. As such, a \emph{valid window} that
contains only and exactly those entailed instances $E^T_{i}$ at every
instant $i$ may be trivially maintained from the streams $P$ and $N$
continuously entailed by some $\Pi^C$.
\end{nestedsection}

>>>>

\begin{nestedsection}{Continuous Datalog}{semantics}
The first requirement of stream reasoning is a well-defined semantics.
that takes into account both the continuous nature of a subset of the
base axioms of a stream reasoning problem, as well as its
instantaneous truth and entailments, in line with the work of the
reasoning community.  However, it must also provide a continuous
interpretation of those instantaneous entailments in order to
integrate with other stream-processing technologies.  More
specifically, an appropriate semantics for stream reasoning on the
Semantic Web must: allow the expression of OWL 2 ontologies to some
degree of expressivity; not sacrifice the precise semantics thereof
whilst incorporating the semantics of stream processing; and allow for
the continuous derivation of entailments in a concise manner.

Continuous Datalog is an extension of positive Datalog with
stream-processing semantics, with a view to accommodating the aims of
the latter within the well-defined model-theoretic semantics of the
former, which are shared with RIF-Core \cite{w3crifbld} and therefore
support reasoning at the expressivity of OWL 2 RL.  This comprises the
extension of Datalog programs to include sources of continuously
generated/reported facts, and the application of sliding window
semantics over those sources, thereby providing for the interpretation
of Continuous Datalog programs as sequences of instantaneous states
that are each valid positive Datalog programs.  As such, these states
each inherit the semantics of positive Datalog.  Furthermore, it
involves the continuous derivation of entailments from this sequence
of instantaneous states as a set of streams of facts; these streams
share a data model with the streams of continuous queries, as defined
by the stream processing community, thereby allowing the integration
of Continuous Datalog programs into stream processing pipelines.

@@

\begin{definition}[Datalog]
\labeldef{continuous datalog: datalog program}

A conventional Datalog program $\Pi = \langle R, F\rangle$ is a tuple
which comprises a finite set of rules $R$ and a finite set of ground
facts $F$. Facts are extensional atomic formulae of first-order logic,
and rules are of the form:

\[\alpha_0 :- \alpha_1 , \ldots , \alpha_n\]
where $\alpha_0$ is an intensional atomic formula and
$\alpha_1 \ldots \alpha_n$ are a mixture of intensional and
extensional atomic formulae.

@@

When considering a time-annotated Datalog program, in which there may be
many distinct instances of a given fact, temporal information regarding when
instances of that fact became true would be lost if such a program were to entail
sets of plain facts. As such, where a fact $f$ would be entailed by some Datalog
program ${\Pi = \langle R, \{ strip(x) : x \in W^{S,x}_i \} \rangle \vdash f}$,
for some window over some stream $S$, an instance ${f | t}$ of $f$ should be
entailed by the time-annotated Datalog program
${\Pi^T = \langle R, W^{S,x}_i \rangle \vdash f | t}$.

@@

It should be noted that \refdef{continuous datalog: justification} of entailment
justification is equally applicable to time-annotated Datalog programs,
simply replacing entailed and justifying facts with fact instances.
Furthermore, each justification for $f$ in some $F^T$ entails a unique
instance ${f | t}$, such that many instances of $f$ may occur in the set
$E^T$ of fact instances entailed by
${\Pi^T = \langle R, W^{S,x}_i \rangle \vDash E^T}$, each with a single
justification.  As such, entailment justification for fact instances may be
expressed as the function ${just : \mathbb{\Pi^T} , \mathbb{F^T} \rightarrow \{ \mathbb{F^T} \}}$,
which maps a fact instance ${f | t}$, entailed by some
${\Pi^T = \langle R, F^T \rangle}$, to the single minimal subset of $F^T$ that
justifies it.

@@

\begin{definition}[Continuous Datalog Program]\label{def:continuous datalog: CDP}
A Continuous Datalog Program $\Pi^C$ is a tuple $\langle R, F, S, r
\rangle$ which comprises a finite set of rules $R$, a finite set of persistent
facts $F$, a stream of facts $S$ and a range $r \in \mathbb{N}$ for
the sliding window over $S$. This specifies that the streamed instances from $S$ are
to be considered valid from their timestamp for the duration represented
by $r$.

@@

\begin{definition}[State of a $\Pi^C$]\label{def:continuous datalog: CDPt}
The state of a $\Pi^C$ at a given instant $i$ is the time-annotated
Datalog program ${\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle}$,
where ${F^0 = \{ f | 0 : f \in F \}}$ is the set of persistent facts $F$
annotated as becoming true at the first recognised instant.

As both $F$ and ${\{ strip(x) : x \in W^{S,r}_{i} \}}$ are sets of Datalog
axioms, the tuple ${\langle R, F \cup \{ strip(x) : x \in W^{S,r}_{i} \}}$
constitutes a valid Datalog program that may be evaluated by the truth
function of Datalog $T^{D}$, the result of which is the definition for the
instantaneous truth of $\Pi^C$ at instant $i$ by $T^{CD}$.
\begin{multline*}
T^{CD} \left( \langle R, F, S, r \rangle, i \right) = \\
	T^{D} \left( \langle R, F \cup \{ strip(x) : x \in W^{S,r}_{i} \} \right)
\end{multline*}

The instantaneous entailments of a $\Pi^C_i$ are the union of the set of
persistent entailments of ${\langle R, F^0 \rangle \vDash E^0}$ (which are
guaranteed to be entailed by every state, as positive Datalog is monotonic),
and the set $E^T_{i}$ of transient entailments whose justifications are at
least partially contained in $W^{S,r}_i$.
\begin{equation*}
\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle \vDash E^0 \cup E^T_{i}
\end{equation*}
$E^T_i$ is defined more formally as
${E^T_{i} = \{ x : \left( \Pi^C_i \vdash x \right) \wedge
\left( just(\Pi^C_i,x) \cap W^{S,r}_i \neq \varnothing \right) \}}$.
By this definition, $E^0$ is guaranteed to be disjoint from every $E^T_i$
for every state $\Pi^C_i$ of every Continuous Datalog Program $\Pi^C$.
\end{definition}

@@

NEEDS WORK\\
--------------------------------\\
From this, we can see that each entailed instance in some
$E^T_{i}$ may take the same form as an instance of any other fact in a
stream of facts, being $\langle s, p, o \rangle | t$, where $t$ is
the real time associated with the instant $i_{e}$ in which
$\langle s, p, o \rangle$ becomes newly justified by the ${CDP}$ (more
specifically, by some subset of the instances in $W^{S,r}_{i_{e}}$).
As such, the stream of entailment instances $P$ of a ${CDP}$ takes
the same form as the \emph{streams of facts} already described, thus
producing a concise stream of entailments that may be read by
further stream processing.  In addition, the stream of negations $N$
of a ${CDP}$ also takes this form, but differs from the streams
discussed thus far in that it is not ordered by the entailment time
$i_{e}$ of the entailment instances but by the time at which their
justification \emph{ceases} to be valid (i.e. the first instant
${i_{n} \geq i_{e}}$ in which $W^{S,r}_{i_{n}}$ no longer contains
the subset of fact instances first present in $W^{S,r}_{i_{e}}$ that
justified the entailment instance $\langle s, p, o \rangle | t$).
\\--------------------------------

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.  However, it can be shown from these axioms, as in
\ref{sec:semantics: proof for entailments}, that it is also possible
to reconstruct exactly the instantaneous set $E^T_{i}$ of instances
entailed at any given instance $i$ by any $\Pi^C = \{ R, F, S, r \}$
from the sliding windows of sufficient range over its entailed streams
$P$ and $N$ at $i$:
\[ E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i} \]
where ${y \geq x \geq r}$. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^C_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{correct} set $E^T_{i}$. As such, a \emph{valid window} that
contains only and exactly those entailed instances $E^T_{i}$ at every
instant $i$ may be trivially maintained from the streams $P$ and $N$
continuously entailed by some $\Pi^C$.
\end{nestedsection}
